\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{geometry}
\geometry{margin=1in}

\title{Programming Assignment 1: Real vs. Fake Facts About Cities}
\author{Hamza Rashid \\
COMP 550, Fall 2024}
\date{Due: September 27th, 2024}

\begin{document}

\maketitle

\section{Problem Setup}
The aim of this assignment is to test the efficacy of linear classifiers
in distinguishing real from fake facts about cities, and determine if the choice
of classifier is non-trivial. In particular, we evaluate the performance 
of a Linear Support Vector Machine, Multinomial Naive Bayes Classifier, 
Logistic Regression model, and Linear Regression model. Lastly, we build a training pipeline
to evaluate the performance of these classifiers under different preprocessing 
techniques.

% We hypothesize that linear models can adequately perform this task
% given an simple dataset with non-overlapping patterns between the 
% two classes, but struggle when patterns frequently overlap.

\section{Dataset Generation and Procedure}
Using an online AI tool, ChatGPT, we generated real and fake facts about Saint Petersburg, Russia. Three prompts were used:
\begin{itemize}
    \item "I would like to create a dataset containing facts about cities. 
    I am particularly interested in Saint Petersburg, Russia.
    List 100 verified facts about Saint Petersburg, Russia."
    \item "now the reason i am asking for this data is to conduct a simple NLP project. 
    As such, for pedagogical purposes, can you generate 100 fake facts about Saint Petersburg, Russia."
    \item (to match the uniform distribution of data provided by another student) "Give me 150 more real and fake facts"
\end{itemize}
We created two files: \texttt{facts.txt} (real facts) 
and \texttt{fakes.txt} (fake facts), each containing 2,750 samples, 
uniformly distributed across 11 cities. The dataset was divided into 
training, validation, and testing sets
with a 70/10/20 split. We utilized sklearn's relative term frequency
vectorizer (\texttt{tfidf}), with inverse document frequency 
(penalizing words that appear in many documents), 
and enabled to detect unigrams, bigrams, and trigrams (city names, and names of historical 
events tend to be split into 1-3 words). We created a data processing pipeline
that culminated in the use of three major preprocessing techniques
tested separately on the classifiers: 
Porter-Stemmer, 
Lemmatization (NLTK, WordNet database), 
and supplementing the WordNet Lemmatizer 
with Part of Speech tagging 
to improve its accuracy.
Minor preprocessing decisions applied 
to each of the aforementioned techniques
included: converting text to lowercase, and
removing all but alphanumeric and whitespace characters.

\section{Parameter Settings and Models}
We experimented with four linear classifiers: 
Linear Support Vector Machine, 
Multinomial Naive Bayes Classifier, 
Logistic Regression model, 
and Linear Regression model. Due to the 
simplicity of our dataset, we tested the parameters
that address overfitting.
We tested the inverse-regularization parameter
for the SVM and Logistic model with values $\{\texttt{0.25, 0.5, 1.0}\}$.
As for the Naive Bayes classifier, 
we compared Laplace smoothing 
with varying degrees of Lidstone smoothing,
testing values $\{\texttt{0.25, 0.5, 1.0}\}$. 
No hyperparameter testing was performed for the linear regression classifier.
We also experimented with various train, validation, test split ratios.


\section{Results and Conclusions}
While the Logistic model performed best with \texttt{C:=1.0} independently of
the preprocessing method, we obtained mixed results with the SVM;
\texttt{C:=1.0} was favorable under Porter-Stemmer and Lemmatization,
while the POS-supplemented Lemmatizer caused a little bit more overfitting,
with \texttt{C:=0.5} being the optimal setting. 
As for the smoothing parameter in the Naive Bayes classifier, 
we did not find any noticeable differences across the tested values.
However, these results were obtained under the 70/10/20 split, 
and while it is expected, it is important to note that 
the optimal parameter settings and overall model performance 
also depend on how the dataset split.

All but the Linear Regression classifier
performed similarly across the major preprocessing methods 
and hyperparameter settings.
% One would expect improvement in accuracy going from 
% Porter-Stemmer to Lemmatization 
% (as it is context-sensitive),
% and after supplementing \texttt{NLTK's WordNet Lemmatizer} 
% with the Part of Speech tagger \texttt{pos\_tag}
% (as the Lemmatizer would be more accurate). 
The relatively 
poor $R^2$ score of  \texttt{$\sim$ 86\%} from the linear regression classifier
may be attributed to a feature vector's tendency to join
a cluster of other data points in its class, due to the 
simplicity of our dataset.
% (for example, catergorizing someone as short or tall can 
% be modelled more accurately by a linear regression model).
On the other hand,
the high accuracy scores, in the neighbourhood of \texttt{98\%}, across the 
SVM, Naive Bayes Classifier, and Logistic Regression model may be attributed
to such clustering of classes. 
Thus, the choice of linear classifier is non-trivial, as its
performance depends
on whether the data points are clustered 
or linearly distributed according to some input,
and if the prediction space is continuous or discrete.


\section{Limitations and Generalization}
The primary limitation of this study is the simplicity of the dataset.
The \texttt{fake} documents use a particular set of words 
frequently (e.g., 'secret', 'underground', 'rumored')
that do not appear as much in the counterpart class, and
tend to be shorter in length than \texttt{fact} documents.
As a result, the classes tend to form seperate clusters of feature vectors,
explaining the high accuracy of all the classifiers we tested. However,
fake facts can utilize the prominent patterns
of a real fact to make a claim that can only be proved 
false with prior knowledge on the given subject – often the case when working with real-world data.
Therefore, while it was reasonable to assume that ChatGPT would provide real facts,
it was not reasonable to assume that the generated 
fake facts would be representative of the false
information that occurs in a typical real-world dataset. 
Therefore, we cannot generalize the results of this
study to the real world; 
the overall problem of distinguishing real from fake
facts is not necessarly simple enough to be modelled by
a linear classifier.

\end{document}
